---
title: "Max Data"
format: 
    html:
        self-contained: true
---

```{r}
library(tidyverse)
library(readit)
library(here)
data1 <- readit(here("drafts","dataset 1..xlsx"))
data2 <- readit(
    here("drafts", "dataset 2..xlsx"),
    sheet = "Results")
```


You normally don't need all the print statements below but this should make it easier to follow what the code is doing.

- `pivot_longer" turns the data in a longer format. You see that you suddenly have a lot more observations. Basically there is one row for each firm, each observation of total assets, equity, and number of employees.
- You need to know this trick but I am basically trying to extract (`str_extract`) the part of the name variable that consists of 20 followed by two other numbers. Name is created by `pivot_longer` and contains the original varialbe names. Than I make that a number instead of a string.
- I create a nicer variable name from this same `name` variable.
-  `pivot_wider` than creates a new dataset with a row for each company and year and the tree variables that you have in your dataset.

```{r}
data2_tidy <- data2 %>% 
pivot_longer(
    cols = matches("^(Total assets)|(Equity)|(Number)")) %>%
print() %>%
mutate(year = str_extract(name, "20[0-9][0-9]")) %>%
mutate(year = as.numeric(year), value = as.numeric(value)) %>%
print() %>%
mutate(variable = case_when(
    str_detect(name, "^Total") ~ "total_assets",
    str_detect(name, "^Equity") ~ "equity",
    str_detect(name, "^Number") ~ "employees"
)) %>%
rename(index = `...1`, 
    company_name = `Company name Latin alphabet`) %>%
select(-name) %>%
print() %>%
pivot_wider(id_cols = c(index, company_name, 
    Specialisation, year), 
    names_from = variable, 
    values_from = value)
```

Unfortunately, I don't know how to match the companies to each other. Do you want to do it based on the name? Do you want to use this index that I created and is it the same as the code in the first dataset?