---
author: Stijn Masschelein
title: introduction to Rstudio for Accounting and Finance
---


# Get the data

We establish a connection with the WRDS databases with my account.

```{r wrds_connection, eval = FALSE}
library(here) # To better deal with the directory structure
i_am("auxilary/introduction_to_rstudio_for_accfin.Rmd")
library(RPostgres)
wrds <- dbConnect(Postgres(),
              host='wrds-pgdata.wharton.upenn.edu',
              port=9737,
              dbname='wrds',
              user='stimas',
              sslmode='require')
```

With an SQL string, we define which data we want. We send the SQL query to the database via de connection. We download the actual data and save in an R object. We close the data, see whether our dataset is the right one, and then save the R object as a file on our computer.

```{r wrds_comp, eval = FALSE}
library(tidyverse)
sql1 <- "SELECT year, gvkey, cusip,
       exec_fullname, coname, ceoann,
       execid, bonus, salary,
       stock_awards_fv, stock_unvest_val,
       eip_unearn_num, eip_unearn_val,
       option_awards, option_awards_blk_value,
       option_awards_num, tdc1, tdc2,
       shrown_tot_pct,
       becameceo, joined_co, reason
       FROM EXECCOMP.ANNCOMP
       WHERE year > 2010 AND ceoann = 'CEO'"
res <- dbSendQuery(wrds, sql1)
compensation <- as_tibble(dbFetch(res)) %>%
  rename_all(tolower)
dbClearResult(res)
glimpse(compensation)

saveRDS(compensation, here("data", "us-compensation-new.RDS"))
```

Now, we do the same for the market value of the company.

```{r wrds_value, eval = FALSE}
gvkeys <- unique(compensation$gvkey)
length(gvkeys)
sql2 <- paste0("SELECT fyear, gvkey, mkvalt, ni ",
            "FROM COMP.FUNDA ",
            "WHERE fyear > 2010 ",
            "AND gvkey in ('",
            paste(gvkeys, collapse = "','"),
            "')"
           )
res <- dbSendQuery(wrds, sql2)
value <- as_tibble(dbFetch(res)) %>%
  rename_all(tolower)
dbClearResult(res)
glimpse(value)
value <- distinct(value)
glimpse(value)

saveRDS(value, here("data", "us-value-new.RDS"))
```

# Explore the data

## With the tidyverse

```{r, packages}
library(tidyverse)
```

```{r}
us_comp <- readRDS(here("data", "us-compensation-new.RDS"))
glimpse(us_comp)
```

```{r}
filter(us_comp, gvkey == "001004")
```

```{r}
filter(us_comp, bonus == 0, year == 2012)
```

## A Gamestop to introduce the pipe

1. Look for any company where "gamestop" is in the name of the company.
2. Look for the `gvkey` of Gamestop. Then use `filter` with exactly that key.

```{r gamestop}
filter(us_comp, str_detect(tolower(coname),
                           "gamestop"))
filter(us_comp, gvkey == "145049")
```

## Moving on

```{r}
select(us_comp, year, coname, bonus, salary, total = tdc1)
```

We first build up the dataset with `select` and `mutate` and the pipe `%>%`. When we are satisfied with the result, we can save the dataset as an R object with the name `us_comp_small`.

```{r}
us_comp_small <-
  select(us_comp, year, coname, bonus,
       salary, total = tdc1) %>%
    mutate(salary_percentage = salary/total)
```


## Quick descriptive statistics

```{r}
group_by(us_comp, gvkey) %>%
    summarise(N = n(), N_CEO = n_distinct(execid),
              average = mean(salary),
              sd = sd(salary),
              med = median(salary),
              min = min(salary),
              max = max(salary)) %>%
    ungroup() %>%
  filter(med > 1000)
```
