---
title: Assignment
---

# Setup

```{r}
#| label: setup
library(tidyverse)
library(lubridate)
library(cowplot)
theme_set(theme_cowplot(font_size = 18))
library(here)
i_am("freaky_friday/assignment_solution.qmd")
```

Let's restrict the data to two years. So that the computations do not take too long.

```{r}
#| label: date-parameters
begin <- ymd("2003-01-01")
end <- ymd("2005-12-31")
```

I keep the full stock price data for now.

```{r}
#| label: data
earn_ann <- readRDS(here("data", "freaky_friday", "earn_ann.RDS")) %>%
  filter(anndat >= begin, anndat <= end)
analyst <- readRDS(here("data", "freaky_friday", "analyst.RDS")) %>%
  filter(anndat >= begin, anndat <= end)
all_stocks <- readRDS(here("data", "freaky_friday", "all_stocks.RDS"))
famafrench <- read_csv(file = here("data", "F-F_Research_Data_Factors_daily.csv"),
                       col_names = c("date", "mkt_rf", "smb", "hml", "rf"),
                       skip = 5, col_type = "ddddd") %>%
  mutate(date = ymd(date)) %>%
  mutate_if(is.numeric, ~ . / 100) %>%
  mutate(mkt = mkt_rf + rf) %>%
  print()
```

# Three-Factor model

```{r}
#| label: three-factor-estimate
create_coefs_3factor <- function(n = 6){
  earn_ann %>% head(n = n) %>%
    distinct(permno, anndat) %>%
    mutate(start = anndat - 300, end = anndat - 46) %>%
    left_join(select(all_stocks, permno, date, ret),
              by = join_by(permno == permno, start <= date, end >= date)) %>%
    # Change this line
    left_join(select(famafrench, mkt_rf, rf, smb, hml, date),
              by = join_by(date == date)) %>%
    filter(!is.na(ret), !is.infinite(ret)) %>%
    # Changed this line
    summarise(y = list(cbind(ret - rf)),
              X = list(cbind(alpha = 1, beta_mkt = mkt_rf,
                             beta_size = smb, beta_value = hml)),
              .by = c(permno, anndat)) %>%
    mutate(coefs = pmap(list(X, y), ~ lm.fit(..1, ..2) %>% coef()),
           .by = c(permno, anndat)) %>%
    select(-y, -X)
}

test <- create_coefs_3factor()

microbenchmark::microbenchmark(
                  create_coefs_3factor(1000),
                  times = 10)
```

If it takes on average (noticeably) more than 10 seconds to run the three-factor model on 1000 observations, it is ok to limit the sample to 1 year. You can change the `start` and `end` date above.

# Abnormal Returns

```{r}
#| label: abnormal-returns
N <- nrow(earn_ann)
# Change this line
results <- create_coefs_3factor(N)

abnormal <- results %>%
  unnest_wider(coefs) %>%
  mutate(date75 = anndat + 75) %>%
  left_join(select(all_stocks, permno, date, ret),
            by = join_by(permno == permno,
                         date75 >= date, anndat <= date)) %>% print %>%
  # change this line
  left_join(select(famafrench, mkt_rf, smb, hml, date),
            by = join_by(date == date)) %>%
  mutate(ar = ret - alpha - beta_mkt * mkt_rf - beta_size * smb - beta_value * hml,
         time_frame = if_else(date - anndat <= 1, "short", "long")) %>%
    summarise(car_sum = sum(ar),
              car_prod = prod(1 + ar) - 1,
            .by = c(permno, anndat, time_frame)) %>%
  filter(!is.na(car_sum), !is.na(car_prod)) %>%
  pivot_wider(values_from = c(car_sum, car_prod), names_from = time_frame)
glimpse(abnormal)
```

# Putting it all together

```{r}
#| label: clean-stock-prices
clean_prices <- all_stocks %>%
  filter(prc > 0) %>%
  select(permno, date, prc, shrout) %>%
  mutate(market_value = prc * shrout) %>%
  select(-shrout) %>%
  print()
```

```{r}
#| label: merging-everything
surprise <- earn_ann %>%
  left_join(analyst,
            by = join_by(ticker, anndat, actual, pdf)) %>%
  left_join(abnormal,
            by = join_by(permno, anndat)) %>%
  mutate(date_minus5 = anndat - 5) %>%
  left_join(clean_prices,
            by = join_by(permno, closest(date_minus5 >= date))) %>%
  mutate(surprise = (actual - median) / prc) 
glimpse(surprise)
```

```{r}
#| label: data-cleaning
winsorise <- 5/10000
main <- surprise %>%
  filter(!is.na(surprise)) %>%
  filter(abs(median) < prc, abs(actual) < prc) %>%
  filter(prc > 2) %>%
  mutate(weekday = wday(anndat, label = TRUE)) %>%
  filter(! weekday %in% c("Sat", "Sun")) %>%
  filter(percent_rank(car_sum_long) >= winsorise,
         percent_rank(car_sum_long) <= 1 - winsorise,
         percent_rank(car_sum_short) >= winsorise,
         percent_rank(car_sum_short) <= 1 - winsorise) %>%
  mutate(group = if_else(weekday == "Fri", "Friday", "Non-Friday"),
         year = year(anndat))
```

# Descriptive Plot

```{r}
#| label: quantiles
quantiles <- main %>%
  mutate(sign = case_when(surprise > 0 ~ "positive",
                          surprise < 0 ~ "negative",
                          surprise == 0 ~ "zero")) %>%
  mutate(
    quintile = ntile(surprise, 5),
    .by = c(sign, year)) %>%
  mutate(
    quantile = case_when(sign == "positive" ~ 6 + quintile,
                         sign == "negative" ~ quintile,
                         sign == "zero" ~ 6
                         )
  ) %>%
  glimpse()
```

```{r}
#| label: descriptive-plot
ggplot(quantiles,
       aes(y = car_prod_short, x = quantile, group = group, colour = group)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2) +
  stat_summary(fun = mean, geom = "line") +
  scale_color_grey()
```

```{r}
#| label: all-plots
quantiles %>%
  pivot_longer(c(car_sum_short, car_prod_short, car_sum_long, car_prod_long),
               values_to = "car", names_to = "window") %>%
  ggplot(aes(y = car, x = quantile, group = group, colour = group)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2) +
  stat_summary(fun = mean, geom = "line") +
  scale_color_grey() +
  facet_wrap(~ window, nrow = 2)
```
